Type theories as quotient inductive-recursive types

### 2025-02-12 tex/main.tex
Some points discussed during the meeting and revisions

A typical workaround in practice (as adopted by Kaposi et al.~\cite{Kaposi2019}) is to add equality constructors as rewrite rules after an inductive definition and further identify equality constructors with the reflxivity proof.
%
As the host theory is assumed to have the uniqueness of identity proof and function extensionality, this workaround is anticipatively justified by the fact~\cite{Hofmann1996,Winterhalter2019,Kapulkin2025} that every well-typed term in extensional type theory can be translated to a term without them in intensional type theory.
%
However, this workaround is computationally ill-behaved, as confluence does not hold any more.
\LT[noinline]{Do they actually justify this workaround anywhere?}
%
Any method that reduces the friction introduced by the use of equality constructors without breaking computational properties seem worth exploring.
%

@@

\paragraph*{Contributions}
\begin{itemize}
  \item Exploration of the use of quotient inductive-inductive-recursive types and definitions by overlapping patterns.
    In particular, we give in a type theory:
    \begin{itemize}
      \item a definition of parallel substitution calculus;
      \item a definition of type theory with a universe and $\Pi$-types,
    \end{itemize}
   using quotient inductive-inductive-recursive types and definitions by overlapping patterns to define type substitution and term substitution partially.
  \item Comparison with other definitions using quotient inductive-inductive types.
\end{itemize}

@@

\subsection{Definitions by overlapping patterns} \label{sec:meta:overlapping}
Functions are typically defined informally using pattern matching for convenience and readability. Formally, however, they are either defined using an explicit pattern-matching construct or elaborated into case trees, which correspond to the \emph{eliminator}, following \emph{first-match} semantics~\cite{Cockx2020a} and employing termination and completeness checking.
This elaboration enables conventional notation while making computational behaviour opaque and favouring asymmetric definitions.
For example, the following two definitions are essentially the same

\begin{minipage}[t]{.5\textwidth}
  \begin{alignat*}{5}
    \blank  & + \blank     && : \N \to \N \to \N \\
    0       &  + m         && = m  \\
  n         &  + 0         && = m  \\
  (\suc\;n) &  + m         && = \suc\;(n + m) \\
  n         &  + (\suc\;m) && = \suc\;(n + m)
\end{alignat*}
\end{minipage}
\begin{minipage}[t]{.5\textwidth}
\begin{alignat*}{5}
  \blank         & +'\blank     && : \N \to \N \to \N \\
  0              & +' m         && \defeq m  \\
  (\suc\;n)      & +' 0         && \defeq \suc\,n \\
  (\suc\;n)      & +' (\suc\;m) && \defeq \suc\;(n +' m)
\end{alignat*}
\end{minipage}
\vskip\belowdisplayskip\noindent
after elaboration in \Agda: each clause of $\blank +' \blank$ is strict (hence the symbol $\defeq$ is used) while clauses of $\blank+\blank$ are not necessarily strict by definition.
This semantics for function definitions affects how definitions reduce and therefore how identities are proved.

Cockx explores an alternative semantics for pattern matching~\cite{Cockx2014}, where all clauses are strict, even in the presence of \emph{overlapping patterns}.
Under this semantics, each clause of $\blank + \blank$ is strictly equal to its right-hand side, allowing the symmetry of $\blank + \blank$ to be proved commutative directly by induction on $n$ and $m$.
Thus, when a function $f$ is defined with overlapping patterns, we use the notation $f\;\overrightarrow{p_i} \reduce t_i$ to emphasize its reduction behaviour.

A key requirement for definitions with overlapping patterns is that, in addition to termination and completeness, any clause must eventually compute to the same term, i.e.\ $f$ is confluent.
Cockx shows~\cite[Proposition~2]{Cockx2014} that if a definition of $f$ is
\begin{enumerate}
  \item \emph{structurally recursive}, i.e.\ arguments of recursive calls are structurally smaller, and
  \item \emph{locally confluent}, i.e.\ for any two function clauses $f\;\overrightarrow{p_1} \reduce t_1$ and $f\;\overrightarrow{p_2} \reduce t_2$ with a most general unifer $\sigma$ for $\overrightarrow{p_1}$ and $\overrightarrow{p_2}$, terms $\sigma t_1$ and $\sigma t_2$ have the same normal form, 
\end{enumerate}
then $f$ is confluent---essentially a variant of Newman's Lemma~\cite{Newman1942}.
We call a diagram $\sigma t_2 \Leftarrow f\;\sigma\overrightarrow{p_1} \Rightarrow \sigma t_1$ a \emph{local peak} informally.

\subsection{Overlapping patterns via adding rewrite rules}
Unfortunately, definitions by overlapping patterns are only implemented in an outdated experimental fork of \Agda.
To emulate a function by overlapping patterns in \Agda, we declare each function clause an additional rewrite rules \cite{Cockx2020,Cockx2021} (via the option \texttt{-{}-rewriting}) and use the local confluence checker (via \texttt{-{}-local-confluence-check}).
This is indeed one of the main use cases of user-defined rewrite rules discussed by Cockx~\cite{Cockx2020}. 
The caveat, though, is that the termination checker is not imposed, so we have to check the termination informally.

@@

\begin{proposition}[Termination]
  Type substitution $[\blank]_{\Ty}\blank$ is terminating.
  \danger
\end{proposition}
\begin{proof}
  Clearly terms on the RHS are structurally smaller~\cite{Abel2002}.
\end{proof}
\begin{proposition}[Local confluence] \label{prop:local-confluence-1}
  Type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
  \danger
\end{proposition}
\begin{proof}
  To show the local confluence, we consider all local peaks, i.e.\ each combination of \eqref{eq:def-type-subst-5} with \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4},
  \[
    [\idS]_{\Ty}\;\UU, \qquad [\sigma;\tau]_{\Ty}\;\UU, \qquad [\pi_1(\sigma, t)]_{\Ty}\;\UU, \qquad\text{and}\qquad [\pi_1(\sigma; \tau)]_{\Ty}\;\UU
  \]
  and show that each applicable clause does result in $\UU$.

@@

As we are defining a function on a quotient inductive type, we have to prove that $[\blank]_{\Ty}\blank$ is coherent, i.e., that it respects all equality constructors.
\LT[noinline]{Should we mention any general condition for coherence?}
\Fredrik[noinline]{Is there one?}
\LT[noinline]{I think Cubical Agda has sort of conditions? (Cubical Agda does not support rewriting)}
\begin{proposition}[Coherence]\label{prop:coherence-1}
  $[\sigma]_{\Ty}\,A$ is (propositionally) equal to $[\tau]_{\Ty}\,A$ for every equality constructor $p : \sigma = \tau$.
  \danger
\end{proposition}
\begin{proof}
  As equality constructors $[\idS]_T$, $[;]_T$, and $[]\UU$ for types become strict, it suffices to show that the following identities
  \begin{align*}
    [\sigma;\idS]_{\Ty} \;A          & = [\sigma]_{\Ty}\;A,
                                     & [\idS;\sigma]_{\Ty} \;A          & = [\sigma]_{\Ty}\;A,
                                     & [(\sigma;\tau);\gamma]_{\Ty} \;A & = [\sigma; (\tau; \gamma)]_{\Ty}A, \\
    [\pi_1(\sigma, t)]_{\Ty}\;A      & = [\sigma]_{\Ty}\;A, \\
    [\sigma; (\tau, t)]_{\Ty} \;A    & = [(\sigma;\tau), [\sigma]\, t]_{\Ty}\;A,
                                     & [\sigma]_{\Ty} \;A               & = [\emptyctx]_{\Ty}\;A,
                                     & [\sigma]_{\Ty} \;A               & = [\pi_1\sigma, \pi_2\sigma]_{\Ty}\;A
  \end{align*}
  hold propositionally.
  The first two identities of the first row hold strictly by \eqref{eq:def-type-subst-1} and \eqref{eq:def-type-subst-2}, while terms of the both sides of the third reduce to $[\sigma]_{\Ty}\;([\tau]_{\Ty}\;([\gamma]_{\Ty}\;A))$ by \eqref{eq:def-type-subst-2}.

@@

  We show the remaining cases by induction on $A$.
  However, $\UU$ is the only constructor for $\Ty\;\Gamma$ for any $\Gamma$ and each of the identities with $A \defeq \UU$ reduce to $\UU$.
\end{proof}
\begin{remark}\label{re:coherence-proof}
  Note that in the above proof the first four identities hold strictly by \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4}, while the last three identities follow from \eqref{eq:def-type-subst-5} by induction on $A$.
  As we extend substitution calculus with other type formers, the first four identities will always hold but the last three depend on how congruence rules for type substitution is defined.
\end{remark}

\LT{We may wonder if the last three identities can be made strict.}

@@

Then, the equality constructor $[]\El$ is accordingly changed to 
\begin{align}
  [ \sigma ]_{\Ty}\,(\El\, u) & \reduce \El\,([\sigma]_{\Tm} {u}) \label{eq:def-type-subst-7}
\end{align}
to retain the local confluence for type substitution:
\begin{proposition}[Termination]
  Type substitution $[\blank]_{\Ty}\blank$ is terminating.
  \danger
\end{proposition}
\begin{proof}
  Again, each recursive call is on a structurally smaller term~\cite{Abel2002}.
\end{proof}
\begin{proposition}[Local confluence] \label{prop:local-confluence-2}
  Type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
  \danger
\end{proposition}
\begin{proof}
  In addition to previous cases in the proof of \cref{prop:local-confluence-1}, consider terms 
  \[
    [\idS]_{\Ty}\;(\El\,u), \qquad [\sigma;\tau]_{\Ty}\;(\El\,u), \qquad [\pi_1(\sigma, t)]_{\Ty}\;(\El\,u), \qquad\text{and}\qquad [\pi_1(\sigma; \tau)]_{\Ty}\;(\El\,u)
  \]
  which reduce to the following terms
  \[
    \El\,u, \qquad \El([\sigma]_{\Tm}\,[\tau]_{\Tm}\,u), \qquad \El([\sigma]_{\Tm}\,u), \qquad\text{and}\qquad
    \El([\sigma]_{\Tm}\,[\pi_1\tau]_{\Tm}\,u)
  \]
  respectively, no matter which function clause of $[\blank]_{\Ty}\blank$ is used.
  Therefore, type substitution is locally confluent.
\end{proof}

@@

To retain the local confluence, we turn $\sigma \uparrow A$ into a simultaneously defined function:
\begin{alignat}{3}
  \blank\uparrow \blank & : \implicit{\Gamma, \Delta}\; (\sigma : \Sub\,\Gamma,\Delta) (A : \Ty\,\Delta) \to \Sub\,(\Gamma, [\sigma]_{\Ty}\,A)\, (\Delta, A) \label{eq:type-of-lifting} \\
\idS                \uparrow A & \reduce \idS \nonumber \\
(\sigma ; \tau)     \uparrow A & \reduce (\sigma \uparrow \sub{\tau}{A} ) ; (\tau \uparrow A) \nonumber \\
\pi_1(\sigma, t)    \uparrow A & \reduce \sigma \uparrow A \nonumber \\
\pi_1(\sigma; \tau) \uparrow A & \reduce (\sigma \uparrow \sub{\pi_1 \tau}{A}) ; (\pi_1 \tau \uparrow A)\nonumber \\
\sigma              \uparrow A & \reduce \sigma^+ , \quad \text{otherwise} \nonumber
\end{alignat}
where $\sigma^+ \defeq (\pi_1\idS ; \sigma, \pi_2\idS)$.
Then, we add another clause for type substitution and an equality constructor for term substitution without transport:
\begin{alignat}{5}
                     & [ \sigma ]_{\Ty}\,(\Pi\,A\,B) && \reduce \Pi\,(\sub{\sigma}{A})\,(\sub{\sigma\uparrow A}{B}) \label{eq:def-type-subst-9} \\
  \mathsf{[]\lambda} & : [\sigma]_{\Tm} (\lambda\,t) && =^{\Tm\,\Gamma\,([\sigma]_{\Ty}(\Pi\,A\,B))} && \lambda\,([\sigma \uparrow A]_{\Tm}\,t) \nonumber
\end{alignat}

@@

\begin{proposition} \label{prop:correctness-2}
  For every $\sigma : \Sub\;\Gamma\;\Delta$, and $A : \Ty\;\Delta$, the lifting
  $\sigma \uparrow A$ is propositionally equal to $(\pi_1 \idS ; \sigma , \pi_2 \idS)$.
\end{proposition}
\begin{proof}
  We prove the statement by induction on $\sigma$ with \cref{lem:lifting}.
  Every case follows from equational reasoning straightforwardly.
  For example, if $\sigma$ is a composite $\sigma;\tau$, we have
  \begin{align*}
    (\sigma;\tau) \uparrow A & \judgeq (\sigma \uparrow [\tau]\,A);(\tau \uparrow A) && \text{by definition} \\
                             & = \sigma^+;\tau^+                                     && \text{by induction hypothesis} \\ 
                             & = (\sigma;\tau)^+                                     && \text{by \cref{lem:lifting}.} 
  \end{align*}
  Other cases follow similarly.
\end{proof}

@@

The type of telescopes over a ``base'' context $\Gamma$ $\Tel : \Ctx \to \Set$ are defined inductive-recursively simultaneously with a concatenation $\blank \dplus \blank$ as follows.

@@

\begin{minipage}{.5\textwidth}
\begin{alignat*}{5}
  & [ \idS   ]_{\Tel}       \,\Xi && \reduce \Xi \\
  & [ \sigma ; \tau ]_{\Tel}\,\Xi && \reduce [ \sigma ]_{\Tel}\,([\tau]_{\Tel}\,\Xi)
\end{alignat*}
\end{minipage}
\begin{minipage}{.5\textwidth}
\begin{alignat*}{5}
  & \idS          \upuparrows \Xi && \reduce \idS \\
  & (\sigma;\tau) \upuparrows \Xi && \reduce (\sigma \upuparrows [ \tau ]_{\Tel} \Xi) ; (\tau \upuparrows \Xi)
\end{alignat*}
\end{minipage}\vskip\belowdisplayskip\noindent
whose confluence property is easy to see.

@@

We conclude this extension with $\Pi$-types by noting that for $t : \Tm\,\Gamma\,(\Pi\,A\,B)$ and $u : \Tm\,\Gamma\,A$, also ``ordinary'' application
$t \mathop{\$} u \defeq [ (\idS , t) ]_{\Tm} (\mathsf{app}\,t)$ of type $\Tm\,\Gamma\,([(\idS , t)] B)$ can be defined without any transport. This is in contrast to the pure QII situation, where application is defined as $[ (\idS, \alert{\transfib{\Tm\;\Gamma}{[id]_T^{-1}}{\color{black}t}}) ]_{\Tm} (\mathsf{app}\,t)$.
%
%\begin{alignat*}{3}
%  \left<\blank\right> & : \implicit{\Gamma, A} \Tm\;\Gamma\;A \to \Sub\;\Gamma\;(\Gamma, A) \\
%  \left< t \right>    & \defeq (\idS , t),
%\end{alignat*}
%used to define the ordinal application: $t \mathop{\$} u \defeq [ \left< u \right> ]_{\Tm} (\mathsf{app}\,t)$, 
%can be introduced without any transport, which is $(\idS, \alert{\transfib{\Tm\;\Gamma}{[id]_T^{-1}}{\color{black}t}})$ in its QII definition.
Similarly, the structural rule for substitution on $\mathsf{app}$
\begin{align*}
  [\sigma \uparrow A ]_{\Tm}(\mathsf{app}\;t) & = \mathsf{app}\,\lambda([\sigma \uparrow A ]_{\Tm}(\mathsf{app}\;t)) && \text{by $\Pi\beta^{-1}$} \\
                                              & = \mathsf{app}\,\left([\sigma]\,\lambda\,(\mathsf{app}\,t)\right) && \text{by $[]\lambda^{-1}$} \\
                                              & = \mathsf{app}\,([\sigma]\,t) && \text{by $\Pi\eta^{-1}$.}
\end{align*}
can be derived without any transport (cf.\ \cite{Altenkirch2016a}), since \eqref{eq:def-type-subst-8} has become strict.

@@

\section{The elimination principle and models} \label{sec:models}
\LT{%
\begin{enumerate}
  \item Show that two definitions are equivalent (ongoing).
  \item Show that methods are equivalent (ongoing).
\end{enumerate}
}
With the QIIR definition of a type theory, we show how to \emph{eliminate} $\Ctx$, $\Ty\,\Gamma$, $\Sub\;\Gamma\;\Delta$, and $\Tm\,\Gamma\,A$ to another types by functions subject to equations corresponding to point constructors, equality constructors, and function clauses (\cref{subsec:eliminator}) .
Then, we can give this QIIR definition of type theory a standard model (\cref{subsec:std-model}) and compare our QIIR definition with its QII counterpart syntactically and semantically (\cref{subsec:equivalence}).

@@

In the QII counterpart, none of identities hold strictly, since their indices always involve equality constructors.
E.g., the constructor $[\idS]_{\Ty}$ gives rise to
$[\idS]^M_{\Ty} : [ \idS^M ]^M_{\Ty}\, A^M =^{\Ty^M\;\Gamma^M}_{\alert{[\idS]_{\Ty}}} A^M$
and the transport along $[\idS]_{\Ty}$ is needed, because $[ \idS^M ]^M_{\Ty}\, A^M$ is indexed by $[\idS]\,A$ instead of~$A$.
Accordingly, this identity becomes $\lam{\gamma} A^M\,\gamma =_{\alert{[\idS]}}^{\lam{\nameless}\Gamma^M \to \Set} A^M$ in the standard model and needs one step~\cite[Lemma~2.3.5]{UFP2013} for reasoning about the constant transport to show.

@@

\section{Formal implementation} \label{sec:formal-implementation}
\LT{Mention the NbE for SC here}
\LT{Compare some formal proofs with pen-and-paper proofs here}
\section{Conclusions, and related and future work}

>>>>

A typical workaround in practice \cite{Kaposi2019,Kaposi2024a} is to add equality constructors as rewrite rules after an inductive definition and further identify equality constructors with the reflxivity proof.
%
As the host theory is assumed to satisfy the uniqueness of identity proof and function extensionality, this workaround can be justified by Hofmann's result~\cite{Hofmann1996,Winterhalter2019,Kapulkin2025} that every well-typed term in extensional type theory can be translated to a term without them in intensional type theory.
%
However, this workaround is computationally ill-behaved, as confluence does not hold in general.
Also, the presentation often differs from the actual definition of type theory in intensional type theory.
%
Any method that reduces the friction introduced by the use of equality constructors without compromising on computational properties and the gap between formalisation and presentation seem worth exploring.
%

@@

\paragraph*{Contributions}
\begin{itemize}
  \item Exploration of the use of quotient inductive-inductive-recursive types and definitions by overlapping patterns.
    In particular, we give in a type theory:
    \begin{itemize}
      \item a definition of parallel substitution calculus;
      \item a definition of type theory with a universe and $\Pi$-types,
    \end{itemize}
   with their elimination principles and the standard model, 
   using quotient inductive-inductive-recursive types and definitions by overlapping patterns to define type substitution and term substitution partially.
  \item Comparison with definitions using quotient inductive-inductive types.
\end{itemize}

@@

\subsection{Definitions by overlapping patterns} \label{sec:meta:overlapping}
\begin{figure} 
\begin{minipage}[t]{.5\textwidth}
  \begin{alignat*}{5}
    \blank  & + \blank     && : \N \to \N \to \N \\
    0       &  + m         && = m  \\
  n         &  + 0         && = m  \\
  (\suc\;n) &  + m         && = \suc\;(n + m) \\
  n         &  + (\suc\;m) && = \suc\;(n + m)
\end{alignat*}
\end{minipage}
\begin{minipage}[t]{.5\textwidth}
\begin{alignat*}{5}
  \blank         & +'\blank     && : \N \to \N \to \N \\
  0              & +' m         && \defeq m  \\
  (\suc\;n)      & +' 0         && \defeq \suc\;n \\
  (\suc\;n)      & +' (\suc\;m) && \defeq \suc\;(n +' m)
\end{alignat*}
\end{minipage}
\caption{Two seemingly different definitions of addition}
\label{fig:additions}
\end{figure}
Functions are typically defined informally using pattern matching for convenience and readability. Formally, however, they are either defined using an explicit pattern-matching construct or elaborated into case trees, which correspond to the \emph{eliminator}, following \emph{first-match} semantics~\cite{Cockx2020a} and employing termination and completeness checking.
This elaboration enables conventional notation while making computational behaviour opaque and favouring asymmetric definitions.
For example, two definitions of addition in \cref{fig:additions} are the same
after elaboration in \Agda: eliminating the first argument of $+$ distinguishes $0$ and $\suc\;n$ and refines the second and the third clauses into clauses of $+'$ and makes the fourth clause useless.
Each clause of $+'$ is strict (hence the symbol $\defeq$ is used) while clauses of $+$ are not necessarily strict by definition.
This semantics for function definitions decides how definitions reduce and therefore how identities are proved.

Cockx~\cite{Cockx2014} explores an alternative semantics for pattern matching, where all clauses are strict, even in the presence of \emph{overlapping patterns}.
Thus, when a function $f$ is defined with overlapping patterns, the notation $f\;\overrightarrow{p_i} \reduce t_i$ is used to emphasize its reduction behaviour.
Under this semantics, each clause of $\blank + \blank$ is strictly equal to its right-hand side, allowing the symmetry addition to be proved commutative directly by induction on $n$ and $m$.

A key requirement for definitions with overlapping patterns is that, in addition to termination and completeness, any clause must eventually compute to the same term, i.e.\ $f$ is confluent.
Cockx shows that if a definition of $f$ is
\begin{enumerate}
  \item \emph{structurally recursive}~\cite{Abel2002}, i.e.\ arguments of recursive calls are structurally smaller, and
  \item \emph{locally confluent}, i.e.\ for any two function clauses $f\;\overrightarrow{p_1} \reduce t_1$ and $f\;\overrightarrow{p_2} \reduce t_2$ with a most general unifer $\sigma$ for $\overrightarrow{p_1}$ and $\overrightarrow{p_2}$, terms $\sigma t_1$ and $\sigma t_2$ have the same normal form, 
\end{enumerate}
then $f$ is confluent---essentially a variant of Newman's Lemma~\cite{Newman1942}.
We call a diagram $\sigma t_2 \Leftarrow f\;\sigma\overrightarrow{p_1} \Rightarrow \sigma t_1$ used in the analysis of local confluence a \emph{local peak}.

\subparagraph*{Implementing overlapping patterns via rewrite rules}
Unfortunately, definitions by overlapping patterns are only implemented in an outdated experimental fork of \Agda.
To emulate a function by overlapping patterns in \Agda, we declare each function clause an additional rewrite rules \cite{Cockx2020,Cockx2021} (via the option \texttt{-{}-rewriting}) and use the local confluence checker (via \texttt{-{}-local-confluence-check}).
This is indeed one of the main use cases of user-defined rewrite rules discussed by Cockx~\cite{Cockx2020}. 
The caveat, though, is that the termination checker is not imposed, so we have to check the termination informally.

@@

\begin{proposition}[Termination]
  Type substitution $[\blank]_{\Ty}\blank$ is terminating.
  \danger
\end{proposition}
%\begin{proof}
%  Clearly terms on the RHS are structurally smaller~\cite{Abel2002}.
%\end{proof}
\begin{proposition}[Local confluence] \label{prop:local-confluence-1}
  Type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
  \danger
\end{proposition}
\begin{proof}
  To show the local confluence, we consider all local peaks, i.e.\ each combination of \eqref{eq:def-type-subst-5} with \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4},
  \[
    [\idS]_{\Ty}\;\UU, \qquad [\sigma;\tau]_{\Ty}\;\UU, \qquad [\pi_1(\sigma, t)]_{\Ty}\;\UU, \qquad\text{and}\qquad [\pi_1(\sigma; \tau)]_{\Ty}\;\UU
  \]
  and show that each applicable clause does result in $\UU$.

@@

As we are defining a function on a quotient inductive type, we have to prove that $[\blank]_{\Ty}\blank$ is coherent, i.e., that it respects all equality constructors.
%\LT[noinline]{Should we mention any general condition for coherence?}
%\Fredrik[noinline]{Is there one?}
%\LT[noinline]{I think Cubical Agda has sort of conditions? (Cubical Agda does not support rewriting)}
\begin{proposition}[Coherence]\label{prop:coherence-1}
  $[\sigma]_{\Ty}\,A$ is (propositionally) equal to $[\tau]_{\Ty}\,A$ for every equality constructor $p : \sigma = \tau$.
  \danger
\end{proposition}
\begin{proof}
  As equality constructors $[\idS]_T$, $[;]_T$, and $[]\UU$ for types become strict, it suffices to show that the following identities
  \begin{align*}
    [\sigma;\idS]_{\Ty} \;A          & = [\sigma]_{\Ty}\;A,
                                     & [\idS;\sigma]_{\Ty} \;A          & = [\sigma]_{\Ty}\;A,
                                     & [(\sigma;\tau);\gamma]_{\Ty} \;A & = [\sigma; (\tau; \gamma)]_{\Ty}A, \\
    [\pi_1(\sigma, t)]_{\Ty}\;A      & = [\sigma]_{\Ty}\;A, \\
    [\sigma; (\tau, t)]_{\Ty} \;A    & = [(\sigma;\tau), [\sigma]\, t]_{\Ty}\;A,
                                     & [\sigma]_{\Ty} \;A               & = [\emptyctx]_{\Ty}\;A,
                                     & [\sigma]_{\Ty} \;A               & = [\pi_1\sigma, \pi_2\sigma]_{\Ty}\;A
  \end{align*}
  hold propositionally.
  The first two identities of the first row hold strictly by \eqref{eq:def-type-subst-1} and \eqref{eq:def-type-subst-2}, while terms of the both sides of the third reduce to $[\sigma]_{\Ty}\;([\tau]_{\Ty}\;([\gamma]_{\Ty}\;A))$ by \eqref{eq:def-type-subst-2}.

@@

  We show the remaining cases by induction on $A$.
  However, $\UU$ is the only constructor for $\Ty\;\Gamma$ for any $\Gamma$ and each of the identities with $A \defeq \UU$ reduce to $\UU$.
\end{proof}
\begin{remark}\label{re:coherence-proof}
  In the above proof the first four identities hold strictly by \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4}, while the last three follow from \eqref{eq:def-type-subst-5} by induction on $A$.
  As we extend substitution calculus with other type formers, the first four identities will always hold but the last three depend on how type substitution is defined.
\end{remark}

@@

Then, the equality constructor $[]\El$ is accordingly changed to 
\begin{align}
  [ \sigma ]_{\Ty}\,(\El\, u) & \reduce \El\,([\sigma]_{\Tm} {u}) \label{eq:def-type-subst-7}
\end{align}
to retain the local confluence for type substitution:
\begin{proposition}[Termination]
  Type substitution $[\blank]_{\Ty}\blank$ is terminating.
  \danger
\end{proposition}
%\begin{proof}
%  Again, each recursive call is on a structurally smaller term~\cite{Abel2002}.
%\end{proof}
\begin{proposition}[Local confluence] \label{prop:local-confluence-2}
  Type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
  \danger
\end{proposition}
\begin{proof}
  In addition to previous cases in the proof of \cref{prop:local-confluence-1}, consider terms 
  \[
    [\idS]_{\Ty}\;(\El\,u), \qquad [\sigma;\tau]_{\Ty}\;(\El\,u), \qquad [\pi_1(\sigma, t)]_{\Ty}\;(\El\,u), \qquad\text{and}\qquad [\pi_1(\sigma; \tau)]_{\Ty}\;(\El\,u)
  \]
  which reduce to the following terms
  \[
    \El\,u, \qquad \El([\sigma]_{\Tm}\,[\tau]_{\Tm}\,u), \qquad \El([\sigma]_{\Tm}\,u), \qquad\text{and}\qquad
    \El([\sigma]_{\Tm}\,[\pi_1\tau]_{\Tm}\,u)
  \]
  respectively, no matter which function clause of $[\blank]_{\Ty}\blank$ is used.
  Therefore, type substitution is locally confluent.
\end{proof}

@@

To retain the local confluence, we turn $\sigma \uparrow A$ into a simultaneously defined function:
\begin{alignat}{3}
  \blank\uparrow \blank & : \implicit{\Gamma, \Delta}\; (\sigma : \Sub\,\Gamma,\Delta) (A : \Ty\,\Delta) \to \Sub\,(\Gamma, [\sigma]_{\Ty}\,A)\, (\Delta, A) \label{eq:type-of-lifting} \\
\idS                \uparrow A & \reduce \idS \nonumber \\
(\sigma ; \tau)     \uparrow A & \reduce (\sigma \uparrow \sub{\tau}{A} ) ; (\tau \uparrow A) \nonumber \\
\pi_1(\sigma, t)    \uparrow A & \reduce \sigma \uparrow A \nonumber \\
\pi_1(\sigma; \tau) \uparrow A & \reduce (\sigma \uparrow \sub{\pi_1 \tau}{A}) ; (\pi_1 \tau \uparrow A)\nonumber \\
\sigma              \uparrow A & \reduce \sigma^+ , \quad \text{otherwise} \nonumber
\end{alignat}
where $\sigma^+ \defeq (\pi_1\idS ; \sigma), \pi_2\idS$.
Then, we add another clause for type substitution and an equality constructor for term substitution without transport:
\begin{alignat}{5}
                     & [ \sigma ]_{\Ty}\,(\Pi\,A\,B) && \reduce \Pi\,(\sub{\sigma}{A})\,(\sub{\sigma\uparrow A}{B}) \label{eq:def-type-subst-9} \\
  \mathsf{[]\lambda} & : [\sigma]_{\Tm} (\lambda\,t) && =^{\Tm\,\Gamma\,([\sigma]_{\Ty}(\Pi\,A\,B))} && \lambda\,([\sigma \uparrow A]_{\Tm}\,t) \nonumber
\end{alignat}

@@

\begin{proposition} \label{prop:correctness-2}
  For every $\sigma : \Sub\;\Gamma\;\Delta$, and $A : \Ty\;\Delta$, the lifting
  $\sigma \uparrow A$ is propositionally equal to $\sigma^+$. 
\end{proposition}
\begin{proof}
  We prove the statement by induction on $\sigma$ with \cref{lem:lifting}.
  Every case follows from equational reasoning straightforwardly.
  For example, if $\sigma$ is a composite $\sigma;\tau$, we have
  \begin{align*}
    (\sigma;\tau) \uparrow A & \judgeq (\sigma \uparrow [\tau]\,A);(\tau \uparrow A) && \text{by definition} \\
                             & = \sigma^+;\tau^+                                     && \text{by induction hypothesis} \\ 
                             & = (\sigma;\tau)^+                                     && \text{by \cref{lem:lifting}.} 
  \end{align*}
  Other cases follow similarly.
\end{proof}

@@

The type of telescopes over a `base' context $\Gamma$ $\Tel : \Ctx \to \Set$ are defined inductive-recursively simultaneously with a concatenation $\blank \dplus \blank$ as follows.

@@

\begin{minipage}{.5\textwidth}
\begin{alignat*}{5}
  & [ \idS   ]_{\Tel}       \,&& \Xi && \reduce \Xi \\
  & [ \sigma ; \tau ]_{\Tel}\,&& \Xi && \reduce [ \sigma ]_{\Tel}\,([\tau]_{\Tel}\,\Xi)
\end{alignat*}
\end{minipage}
\begin{minipage}{.5\textwidth}
\begin{alignat*}{5}
  & \idS          && \upuparrows \Xi && \reduce \idS \\
  & (\sigma;\tau) && \upuparrows \Xi && \reduce (\sigma \upuparrows [ \tau ]_{\Tel} \Xi) ; (\tau \upuparrows \Xi)
\end{alignat*}
\end{minipage}\vskip\belowdisplayskip\noindent
whose confluence property is easy to see.

@@

We conclude this extension with $\Pi$-types by noting that for $t : \Tm\,\Gamma\,(\Pi\,A\,B)$ and $u : \Tm\,\Gamma\,A$, also `ordinary' application
$t \mathop{\$} u \defeq [ (\idS , t) ]_{\Tm} (\mathsf{app}\,t)$ of type $\Tm\,\Gamma\,([(\idS , t)] B)$ can be defined without any transport. This is in contrast to the pure QII situation, where application is defined as $[ (\idS, \alert{\transfib{\Tm\;\Gamma}{[id]_T^{-1}}{\color{black}t}}) ]_{\Tm} (\mathsf{app}\,t)$.
%
%\begin{alignat*}{3}
%  \left<\blank\right> & : \implicit{\Gamma, A} \Tm\;\Gamma\;A \to \Sub\;\Gamma\;(\Gamma, A) \\
%  \left< t \right>    & \defeq (\idS , t),
%\end{alignat*}
%used to define the ordinal application: $t \mathop{\$} u \defeq [ \left< u \right> ]_{\Tm} (\mathsf{app}\,t)$, 
%can be introduced without any transport, which is $(\idS, \alert{\transfib{\Tm\;\Gamma}{[id]_T^{-1}}{\color{black}t}})$ in its QII definition.
Similarly, the structural rule for substitution on $\mathsf{app}$
\begin{align*}
  [\sigma \uparrow A ]_{\Tm}(\mathsf{app}\;t) & = \mathsf{app}\,\lambda([\sigma \uparrow A ]_{\Tm}(\mathsf{app}\;t)) && \text{by $\Pi\beta^{-1}$} \\
                                              & = \mathsf{app}\,\left([\sigma]\,\lambda\,(\mathsf{app}\,t)\right) && \text{by $[]\lambda^{-1}$} \\
                                              & = \mathsf{app}\,([\sigma]\,t) && \text{by $\Pi\eta^{-1}$.}
\end{align*}
can be derived without any transport (cf.\ \cite{Altenkirch2016a}), since \eqref{eq:def-type-subst-8} has become strict.

@@

\section{The elimination principle and models} \label{sec:models}
\LT{%
\begin{enumerate}
  \item Show that two definitions are equivalent (ongoing).
  \item Show that methods are equivalent (ongoing).
\end{enumerate}
}
With the QIIR definition of a type theory, we show how $\Ctx$, $\Ty\,\Gamma$, $\Sub\;\Gamma\;\Delta$, and $\Tm\,\Gamma\,A$ are eliminated to another types by functions subject to equations corresponding to point constructors, equality constructors, and function clauses (\cref{subsec:eliminator}).
Then, we can give this QIIR definition of type theory a standard model (\cref{subsec:std-model}) and compare our QIIR definition with its QII counterpart (\cref{subsec:equivalence}).

@@

In the QII counterpart, none of identities hold strictly: their indices always involve equality constructors.
E.g., the constructor $[\idS]_{\Ty}$ gives rise to
$[\idS]^M_{\Ty} : [ \idS^M ]^M_{\Ty}\, A^M =^{\Ty^M\;\Gamma^M}_{\alert{[\idS]_{\Ty}}} A^M$
and the transport along $[\idS]_{\Ty}$ is needed, because $[ \idS^M ]^M_{\Ty}\, A^M$ is indexed by $[\idS]\,A$ instead of~$A$.
Accordingly, this identity becomes $\lam{\gamma} A^M\,\gamma =_{\alert{[\idS]}}^{\lam{\nameless}\Gamma^M \to \Set} A^M$ in the standard model and needs an extra step~\cite[Lemma~2.3.5]{UFP2013} for reasoning about the constant transport.

@@

In general, we can translate the QIIR definition of type theory to its QII definition

\LT{Mention the NbE for SC here}
\section{Formal implementation} \label{sec:formal-implementation}
\LT{Compare some formal proofs with pen-and-paper proofs here}
\section{Conclusions, and related and future work}

We have used quotient inductive-recursive types and overlapping
patterns to represent type theory in type theory. By defining type
substitution recursively with overlapping patterns, we can avoid most
transports that show up in terms in other formalisations of type
theory in (intensional) type theory. In other words, we have shown that so-called
`transport hell' can be minimised when formalising type theory in type
theory this way, without hiding any details in the formalisation.
